\chapter{Literature Review}

\section{Introduction to Routing Algorithms in NoCs}

Routing determines how packets traverse a NoC from source to destination. In a typical 2D mesh topology, each router forwards packets based on a routing function implemented within its routing unit. The primary objectives of routing are to minimize latency, maintain high throughput, and ensure correctness by avoiding deadlock and livelock.

NoC routing algorithms are broadly classified as deterministic or adaptive. Deterministic (static) routing follows fixed, predefined paths that do not depend on runtime conditions. A widely used example is XY routing \cite{Chawade2012}, where packets are first routed along the X dimension and then along the Y dimension. This approach is simple and hardware-efficient, but it cannot respond to traffic imbalance, faults, or localized congestion, which may lead to performance degradation and uneven resource utilization \cite{AROMa2018}.

To address these limitations, adaptive routing algorithms dynamically select paths based on the current state of the network. Routing decisions may incorporate local or global information, including buffer occupancy, link utilization, fault status, or thermal conditions. Depending on their objective and design, adaptive routing strategies can be categorized into several classes. Following the taxonomy presented in \cite{Kondoth2026}, adaptive routing can be broadly grouped into reactive, proactive, and application-specific approaches.

One important class of reactive adaptation is congestion-aware routing, which aims to improve performance by steering traffic away from congested regions of the network. These algorithms use runtime indicators such as queue lengths or buffer occupancy to balance load and reduce packet delay \cite{Kondoth2026}. Another class of reactive routing is fault-tolerant routing, which ensures continued packet delivery in the presence of permanent faults in routers or links. These approaches detect failures and reroute traffic around faulty components, maintaining network functionality after degradation has occurred \cite{Kondoth2026}. In contrast, proactive routing approaches aim to mitigate degradation before it results in failure. Aging-aware routing falls into this category, where routing decisions are influenced by long-term stress indicators such as temperature and switching activity. By distributing traffic more evenly across the network, these methods seek to reduce the rate of wearout mechanisms \cite{Kondoth2026}. In this thesis, we focus on adaptive routing algorithms. Specifically, we review the state of the art in fault-tolerant, congestion-aware, and aging-aware routing, and identify gaps that motivate our proposed wear-aware routing algorithm based on linear approximate reinforcement learning.

\section{Congestion-Aware Routing Algorithms}

A major branch of adaptive routing research in NoCs focuses on congestion awareness. The main objective of these algorithms is to reduce packet latency and improve throughput by steering traffic away from overloaded routers and links \cite{Kondoth2026}. In contrast to deterministic routing methods such as XY routing, congestion-aware approaches exploit runtime information such as buffer occupancy, queue length, or downstream link status to make more informed path selections. The literature in this area generally evolved from lightweight local schemes, to approaches incorporating wider congestion visibility, and finally to learning-based methods that attempt to infer better routing decisions from experience.

DyAD \cite{Hu2004} combines deterministic and adaptive routing in a lightweight manner. Under low-congestion conditions, it behaves like a deterministic routing algorithm, while under congestion it switches to odd-even adaptive routing using congestion flag wires derived from buffer occupancy. DyAD achieved a throughput improvement of about 61.7\% over XY routing and reduced latency by 14\% relative to odd-even routing. However, DyAD relies only on local congestion information. As a result, its decisions may be shortsighted, since a direction that appears uncongested locally may still lead to heavier contention farther downstream. To overcome the limited visibility of purely local schemes \textit{Destination-Based Adaptive Routing} (DBAR) \cite{Ma2011} augments local routing decisions with propagated congestion information from routers lying in the quadrant between the current router and the destination. DBAR improves average saturation throughput by 25.2\% compared with traditional routing. Even so, DBAR still incurs extra state propagation and hardware complexity, and its effectiveness depends on how accurately the propagated congestion information reflects current network conditions. A further refinement of this idea appears in \textit{Regional Congestion Awareness} (RCA) \cite{Gratz2008}. RCA collects congestion information from a larger region of the network and uses that information to guide path selection more intelligently than purely local methods. The survey describes several RCA variants with different scopes of visibility, and reports latency reductions of 16--71\% relative to dimension-order routing. It also highlights a recurring trade-off in congestion-aware routing: the more broadly congestion information is collected, the greater the risk of communication overhead, stale information, and increased router complexity.

While these classical congestion-aware algorithms improved performance, most of them still relied on explicitly designed heuristics. The router designer had to choose what congestion signals to monitor, how far to propagate them, and how to combine them into a routing decision. This heuristic dependence limits adaptability when traffic conditions vary significantly over time. It also motivates the use of reinforcement learning, where routers can learn routing behavior from experience rather than depending entirely on manually crafted decision rules.

An early step in this direction is the RL-based routing-selection framework discussed in the survey under references \cite{DBLP:journals/corr/abs-1908-04484} and \cite{Reza2021}. In this formulation, reinforcement learning is used not to directly choose an output port, but to select among routing strategies such as XY, north-last, and adaptive north-last based on the observed NoC state. The state includes metrics such as injected and received flits, while the reward is tied to throughput. However, the action space remains relatively coarse since the learner chooses among pre-existing routing algorithms rather than learning a more fine-grained router-level policy. A more direct learning-based approach is represented by \textit{DRLAR} \cite{WANG2024110419}, which formulates routing as a multi-objective optimization problem and uses a lightweight distributed deep Q-network deployed at each router. Instead of merely switching among preset algorithms, DRLAR estimates Q-values from local network state and directly learns routing behavior that balances load, congestion, and endpoint isolation. DRLAR reduces average latency by 33.3\% and improves saturation throughput by up to 20\% over approaches such as Footprint routing. Still, deep RL introduces training cost and implementation overhead, which may limit practicality in hardware-constrained NoC routers. Another notable line of work is \textit{HARAQ} \cite{Ebrahimi2012}, a Q-learning-based routing algorithm in which each switch maintains a Q-table containing estimated latency values for reaching destinations through different output channels. Packets are forwarded through the channel with the lowest estimated cost, and the Q-values are updated using observed feedback. This approach results in latency reductions of up to 37\% compared with methods such as DBAR. However, the state-action space grows rapidly with network size and traffic diversity, and convergence can be slow in dynamic environments.

This scalability problem motivates the use of function approximation. \textit{linear approximate RL-based adaptive routing} \cite{LARE2023}, instead of storing a Q-value for every state-action pair, it estimates Q-values using a linear model over selected features. The reported result is a 10.5\% reduction in packet latency compared with DOR and Footprint routing, while significantly reducing memory requirements relative to Q-table-based approaches. This idea preserves the adaptive nature of RL while making implementation more feasible for NoC routers with strict hardware budgets. In that sense, linear approximation forms a bridge between classical heuristic routing and more costly deep RL approaches, and it is especially relevant for architectures where low-overhead distributed decision-making is required.

\section{Fault-Tolerant Routing Algorithms}

Fault-tolerant routing has been a major area of research in NoC reliability because failures in routers or links can disrupt packet delivery, reduce throughput, and in severe cases partition the network. Unlike congestion-aware routing, which mainly targets performance, fault-tolerant routing is primarily concerned with preserving connectivity and ensuring successful packet delivery in the presence of permanent or intermittent faults \cite{Kondoth2026}. In practice, these algorithms attempt to detect faulty components and reroute traffic through healthy regions of the network while maintaining acceptable latency and throughput.

One representative early approach is the \textit{Fault-Tolerant Minimal Routing} (FTMR) \cite{Zhao2017} algorithm. FTMR seeks to preserve minimal routing whenever possible by identifying the shortest viable path around faults. It does this through two procedures, Forward Allowed-Path-Counter (F-APC) and Reverse Allowed-Path-Counter (R-APC), which respectively identify unreachable and useless nodes before routing decisions are made. The importance of FTMR lies in its explicit attempt to retain minimal-path behavior under fault conditions rather than immediately resorting to broad non-minimal detours. However, this benefit comes at the cost of additional computation, since the path analysis must be carried out in both forward and reverse directions for each routing decision. As a result, FTMR is conceptually strong as a shortest-path fault-tolerant method, but less attractive when router simplicity and low overhead are primary design goals.

A more widely cited baseline in the fault-tolerant NoC literature is the \textit{Fault-on-Neighbor} (FoN) \cite{Feng2010} algorithm. FoN improves routing decisions by allowing a router to observe link status within a two-hop neighborhood and use that information to avoid nearby faults. The survey identifies FoN as a common reference point for evaluating later fault-tolerant routing schemes, which reflects its practical importance in the literature. Its appeal comes from the fact that it introduces fault awareness without requiring a fully global view of the network. At the same time, its reliance on local fault information also reveals a recurring limitation of many reactive fault-tolerant methods: a route that appears safe within a two-hop neighborhood may still lead to congestion, dead ends, or repeated rerouting farther downstream.

Another important direction in the literature is represented by the \textit{Logic-Based Distributed Routing} \cite{Flich2008} family, including LBDR, uLBDR, and d2-LBDR. These methods move away from large routing tables and instead use simple logic operations and configurable bits to encode connectivity and fault status. The survey reports that logic-based approaches can reduce area and power relative to table-driven methods, while still allowing routers to reconfigure paths around faults. The later d2-LBDR extension improves this framework further by maintaining the distance to the nearest known failure and incorporating that information into routing decisions. Its weakness, however, is that per-packet logic evaluation and dynamic reconfiguration can still become costly in large and highly dynamic NoC environments. The d2-LBDR framework also led to more advanced reconfiguration-oriented methods such as \textit{TRACK} \cite{JAIN202092}, which dynamically updates routing in response to directional link failures. TRACK achieves 75\% reliability under double-link failures and improves latency and throughput by up to 42\% and 22\%, respectively, compared with existing fault-tolerant approaches. Still, the emphasis on nearby failures means that such methods may not fully capture more distant fault interactions along the route.

More recently, learning-based fault-tolerant routing has been explored as an alternative to fixed rerouting heuristics. The survey highlights a Q-learning-based routing algorithm in \cite{Samala2020}, where routers learn fault-tolerant forwarding choices by rewarding transitions that avoid unnecessary hops and improve packet delivery. The reported packet delivery reaches 95\% in the presence of 10\% faults, suggesting that reinforcement learning can capture useful routing behavior even under degraded network conditions. A further extension in \cite{Samala2022} applies multi-agent reinforcement learning, where routers make decentralized decisions using a two-level Q-table structure. The study reports a 7.1\% improvement in packet delivery over the earlier Q-learning approach. These works are significant because they show that learning can improve adaptability in fault-tolerant routing, especially when fault patterns are irregular or difficult to encode using static rules. However, they also inherit the main limitation of slow convergence and growing state-action complexity, which may reduce effectiveness during early operation or in rapidly changing fault scenarios.

\section{Aging-Aware Routing Algorithms}

Aging-aware routing emerged as a proactive reliability strategy for Network-on-Chip systems, motivated by the observation that traffic distribution directly affects long-term wearout. Unlike fault-tolerant routing, which reacts after routers or links have already failed, aging-aware routing attempts to delay failure by controlling how stress accumulates across the network over time \cite{Kondoth2026}. This is particularly important because aging mechanisms such as EM, HCI, and BTI are strongly influenced by switching activity, duty cycle, and temperature. As a result, heavily utilized routers and links tend to degrade faster than lightly used ones, making routing policy a critical determinant of NoC lifetime. In this body of work, the primary evaluation metric is typically mean time to failure (MTTF), which reflects the expected operational lifetime of the interconnect under a given routing strategy \cite{Kondoth2026}.

One of the clearest representative approaches in this area is the \textit{stress-aware routing} algorithm in \cite{Clotho2015}. This method continuously monitors wearout risk in NoC components and routes traffic away from routers or links that are judged to be under higher stress due to temperature, utilization, or process-related effects. According to the study, the method improves MTTF by an average of 66\% for EM risk and 8\% for HCI risk relative to DOR. This result that lifetime extension can be achieved through traffic redistribution alone, without waiting for failures to appear. At the same time, the method depends on packet headers to disseminate network-state information, which means routing decisions may be affected by stale information when traffic conditions change quickly. In addition, the authors acknowledge that the method introduces significant hardware overhead due to the need for monitoring stress parameters.

\textit{Lifetime-Aware Routing (DPLR)} \cite{Wang2016}, which introduces a lifetime budget for each router. The core idea is that every router has a maximum workload budget over a given period, and routing decisions should be made with respect to how much of that budget has already been consumed. The study reports that DPLR improves MTTF by 20--55\% compared to conventional routing methods. However, it remains dependent on explicit cost modeling and repeated optimization, which can become difficult to sustain efficiently as the network grows or workload conditions become more dynamic. A more optimization-driven line of work is represented by the \textit{dynamic programming (DP)-based aging-aware routing} approach in \cite{Wang2019}. Rather than simply avoiding highly stressed components, this method explicitly computes routes that minimize the maximum aging rate across the network. In other words, it formulates routing as a constrained optimization problem whose objective is to produce a more uniform wear distribution and thereby extend overall NoC lifetime. The study reports that this approach improves MTTF by up to 20\% over existing aging-aware routing techniques. However, like other dynamic programming methods, it can be computationally expensive and may struggle to converge rapidly in dynamic NoC environments where traffic conditions evolve continuously.

Another notable direction is runtime degradation tracking, represented by methods such as \textit{Dynamic Wearout Resilient Routing} (DWRR) \cite{Ancajas2014}. DWRR monitors router degradation through critical-path delay measurements and then broadcasts wearout information across the network so that traffic can be redistributed more evenly. The study reports a 25\% improvement in MTTF over existing fault-tolerant techniques. However, this method introduces communication overhead, and in rapidly changing traffic conditions the propagated degradation information may already be outdated by the time it is used. A related example of this trade-off appears in the \textit{centralized aging table} approach in \cite{Ghaderi2017}. In this method, a centralized structure tracks flit counts and residence times in routers to estimate temperature fluctuations, stress levels, and cumulative age. Routes are then selected according to the minimum cumulative aging cost from source to destination. The study reports a 39\% improvement in degradation rate of aged routers and a 52\% improvement in aging imbalance relative to traditional algorithms. However, it depends on centralized tracking, periodic updates, or globally maintained state, which reduces responsiveness and increases implementation overhead. In the case described in the survey, routing table updates occur only weekly, which may not accurately reflect more immediate changes in traffic-induced wear.

\section{Motivation for Learning-Based Aging-Aware Routing}

The literature on adaptive routing in NoCs shows a clear progression toward more flexible decision-making. In congestion-aware routing, early methods relied on fixed heuristics \cite{Hu2004} \cite{Gratz2008}, while more recent work extended this direction by introducing reinforcement learning and deep reinforcement learning \cite{DBLP:journals/corr/abs-1908-04484} \cite{Reza2021}, allowing routers to adapt routing behavior based on observed network conditions. Similar trends can be observed in fault-tolerant routing, where learning-based methods such as Q-learning and multi-agent reinforcement learning were introduced to improve routing under irregular fault patterns and dynamic network degradation \cite{Samala2020} \cite{Samala2022}. These developments indicate that, in both congestion management and fault tolerance, the field has already begun moving beyond static heuristics toward policies that can be learned from interaction with the network. In contrast, aging-aware routing has not advanced as far in this direction. Existing aging-aware approaches successfully established that routing decisions can influence long-term reliability by redistributing traffic away from highly stressed routers and links \cite{Clotho2015} \cite{Wang2016}. However, these approaches are still largely based on predefined models, hand-crafted cost functions, and explicit optimization procedures. Their adaptability therefore remains limited by the assumptions embedded in those models.

A recurring limitation in prior aging-aware routing research is the dependence on global or semi-global network knowledge. Several methods rely on packet-header-based dissemination of wear information, centralized aging tables, or broadcast-based monitoring of degradation state \cite{Clotho2015} \cite{Ghaderi2017}. While these mechanisms improve visibility into the aging condition of the network, they also introduce non-trivial communication, storage, and control overhead. More importantly, the information used for decision-making may become stale before it is applied \cite{Kondoth2026}. As a result, many of these approaches are proactive in intent, but only partially adaptive in practice. Another important limitation is that most prior aging-aware routing algorithms treat the routing problem as an explicit optimization problem that must be repeatedly solved using predefined criteria \cite{Clotho2015}. This creates two challenges. First, the resulting routing policy is only as good as the chosen heuristic or optimization model. Second, the policy may not remain effective when traffic patterns, stress distributions, or workload characteristics change significantly over long execution periods. In other words, these methods are adaptive only within the boundaries of their manually defined objective functions.

This stands in contrast to learning-based routing methods developed for congestion avoidance and fault tolerance \cite{Reza2021} \cite{Samala2020}. In those domains, reinforcement learning was introduced specifically to reduce dependence on manually engineered routing rules and to allow routing decisions to improve through experience. Q-learning \cite{Watkins1989}, deep Q-networks \cite{Hester2017}, and linear function approximation \cite{LARE2023} were all explored as ways to capture network behavior more flexibly and with different trade-offs in accuracy, scalability, and memory cost. The existence of these methods in adjacent routing domains suggests a natural next step for aging-aware routing. Rather than repeatedly computing routes from explicit aging models alone, routers can instead learn policies that implicitly capture the relationship between traffic distribution, network state, and long-term wear.

The key research gap, therefore, is not whether aging-aware routing is useful, since prior work has already shown that it is. The gap lies in how aging awareness is currently realized. Existing methods remain dominated by heuristic and model-driven formulations that incur hardware overhead, depend on explicit state propagation, and offer limited flexibility under changing operating conditions. Unlike congestion-aware and fault-tolerant routing, aging-aware routing has not yet been sufficiently extended into a lightweight learning-based framework that can adapt online while remaining practical for distributed router-level implementation. This gap motivates this thesis. The central idea is to move aging-aware routing from explicit heuristic optimization toward learning-based adaptation. By formulating routing as a reinforcement learning problem, the router can learn to avoid highly stressed NoC components through interaction with the network rather than by relying solely on predefined cost functions. At the same time, to remain practical for hardware implementation, such a method must avoid the large storage and convergence costs associated with conventional tabular learning. This makes linear approximate reinforcement learning \cite{LARE2023} the perfect candidate, as it offers a balanced compromise between adaptability and implementation efficiency.